Demographic Species Distribution Models

### 2021-05-12 manuscript/ms.Rmd
merge Ben's edits and some of my own

1. Quantifying the distributions and abundances of species is a problem central to ecology. Correlative species distribution models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they have questionable ability to predict how distributions will change in response to environmental change, habitat loss, or human interventions.

2. I propose demographic species distribution models (DSDMs): matrix population models that are statistically fitted to species distribution data, to form a mechanistically-based statistical model for species distributions. This approach differs from existing mechanistic SDMs in that it makes use of commonly available species distribution data (abundance, presence-only, or presence-absence), explicitly models how vital rates such as fecundity and survival vary over space, and enables - but does not require - the incorporation of prior knowledge of a species' ecology.

3. Example applications mapping the Carolina Chickadee in North America, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates. Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.

4. DSDMs can be extended to consider features of population dynamics such as dispersal, density dependence, and biotic interactions. Being specifically designed for efficient statistical inference, they are a promising replacement for correlative SDMs where study aims require the incorporating of more ecological mechanism.

>>>>

1. Quantifying the distributions and abundances of species is a problem central to ecology. Correlative species distribution models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they often perform poorly at predicting how distributions will change in response to environmental change, habitat loss, or human interventions.

2. I propose demographic species distribution models (DSDMs): matrix population models that are statistically fitted to species distribution data to form a mechanistic statistical model for species distributions. This approach differs from existing mechanistic SDMs in that it makes use of commonly available species distribution data (abundance, presence-only, or presence-absence), explicitly models how vital rates such as fecundity and survival vary over space, and enables - but does not require - the incorporation of prior knowledge of a species' ecology.

3. Example applications mapping the Carolina Chickadee in North Ameerica, and a species of Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs and produce plausible spatial estimates of vital rates. Even when DSDMs are provided with little information on the ecology of the target species, predictions of distributions are well-identified. Where there is uncertainty about a species' population dynamics, this uncertainty is reflected in more uncertain estimates of vital rates.

4. DSDMs can be extended to consider features of population dynamics such as dispersal, density dependence, and biotic interactions. Being specifically designed for efficient statistical inference, they are a promising replacement for correlative SDMs where study aims require the incorporation of more ecological mechanism.
